\documentclass[a4paper,12pt]{amsart}

\input{header.tex}

\begin{document}

\thispagestyle{empty}

\section*{NTIN071 A\&G: Tutorial 8 -- Pumping lemma for context-free languages}

\medskip

\subsection*{Teaching goals:} The student is able to

\begin{itemize}\setlength{\itemsep}{0pt}
    \item give a formal statement of the Pumping lemma for context-free languages
    \item explain the proof of the Pumping lemma for context-free languages
    \item apply the Pumping lemma to prove that a given language is not context-free
\end{itemize}


\section*{In-class problems}



\medskip\begin{problem}[Pumping lemma: statement and proof]
    
    \begin{enumerate}[(a)]\setlength\itemsep{6pt}
        \item Formulate the Pumping Lemma for context-free languages (without consulting your notes).
        \item Compare the statement to the version for regular languages. 
        \item Explain the idea behind its proof.
        \item Demonstrate pumping on the language $L=\{ww^R\mid w\in \{a,b\}^*\}$.
    \end{enumerate}

\end{problem}

\medskip\begin{problem}[Pumping lemma: application]

    Decide if the following languages are context-free. Prove that your answer is correct.
    
    \begin{enumerate}[(a)]
        \setlength\itemsep{6pt}
        \item $L=\{0^i1^j0^i\mid i,j\geq 0\}$
        \item $L=\{0^i1^j0^i\mid 0\leq i\leq j\}$
        \item $L=\{0^i1^j2^k\mid 0\leq i\leq j\leq k\}$
        \item $L=\{ww\mid w\in \{0,1\}^*\}$
        \item $L=\{ww^R\mid w\in \{0,1\}^*, |w|_0=|w|_1\}$
        \item $L=\{1^{n^2+n+1}\mid n\geq 0\}$
    \end{enumerate}

\end{problem}


\section*{Extra Practice and Thinking}


\medskip\begin{problem}[Pumping and right-linear grammars]

    Give an alternative proof of the Pumping lemma for regular languages that is based on derivations from a right-linear grammar.

\end{problem}


\medskip\begin{problem}[Pumping linear languages]
    
    Recall that a grammar is \emph{linear}, if it only contains production rules of the form $A\to uBw$ and $A\to w$, where $A,B\in V$ and $u,w\in T^*$.
    
    \begin{enumerate}[(a)]
        \setlength\itemsep{6pt}
        \item Formulate a Pumping lemma for linear languages.
        \item Proof the statement using derivations from a (reduced) linear grammar.
        \item How does the pumping constant $n$ from the lemma relate to a linear grammar for $L$?
        \item Show that the language $L=\{w\in \{0,1\}^*\mid |w|_0=|w|_1\}$ is not linear.
        \item Where does $L$ lie within the Chomsky hierarchy?
    \end{enumerate}
    
\end{problem}


\medskip\begin{problem}[Pumping lemma: application]

    Decide if the following languages are context-free. Prove that your answer is correct.

    \medskip
    
    \begin{enumerate}[(a)]
        \setlength\itemsep{6pt}
        \item $L=\{0^i1^i\mid i\geq 0\}$        
        \item $L=\{0^i1^j0^i\mid 0\leq j\leq i\}$
        \item $L=\{0^i1^i2^i\mid i\geq 0\}$
        \item $L=\{0^{2i}1^{3i}0^i\mid i\geq 0\}$        
        \item $L=\{ww^R\mid w\in \{0,1\}^*\}$        
        \item $L=\{1^{n^2}\mid n\geq 0\}$        
        \item $L=\{1^p\mid p \text{ is a prime}\}$
        \item $L=\{0^i1^{j}\mid 0\leq i\leq j^2\}$
    \end{enumerate}

\end{problem}


\end{document}


