\documentclass[a4paper,12pt]{amsart}

\input{header.tex}

\begin{document}

\thispagestyle{empty}

\section*{NTIN071 A\&G: Tutorial 10 -- Properties of context-free languages, DPDA}

\medskip

\subsection*{Teaching goals:} The student is able to

    \begin{itemize}\setlength{\itemsep}{0pt}
        \item decide whether context-free languages (CFL) are closed under various operations
        \item give a formal definition of a deterministic pushdown automaton (DPDA)
        \item explain the difference between accepting by final state and by empty stack for DPDA, i.e., the classes of languages $\mathrm{L_{DPDA}}$ and $\mathrm{N_{DPDA}}$
        \item explain and illustrate the relationship between $\mathrm{L_{DPDA}}$ and $\mathrm{N_{DPDA}}$ with other classes
        \item decide whether $\mathrm{L_{DPDA}}$ and $\mathrm{N_{DPDA}}$ are closed under various operations
    \end{itemize}


\section*{In-class problems}


\begin{problem}[Non-palindromes]
    Consider the language $L=\{w\in\{a,b\}^*\mid w=w^R\}$.
    \begin{enumerate}[(a)]
        \item Is $L$ context-free?
        \item Is its complement $\overline{L}$ context-free?
    \end{enumerate}
\end{problem}


\begin{problem}[Closure Properties]
    Are (I) $\mathrm{CFL}$, (II) $\mathrm{L_{DPDA}}$, (III) $\mathrm{N_{DPDA}}$ closed under the following operations? Prove or disprove.
    \begin{multicols}{2}
        \begin{enumerate}[(a)]
            \item Intersection
            \item Complement
            \item Intersection with a regular language
            \item Homomorphism
        \end{enumerate}
    \end{multicols}    
\end{problem}

\begin{problem}
    Show that:
    \begin{enumerate}[(a)]
        \item $\{a^nb^{n+1}\mid n\geq 0\}\in\mathrm{N_{DPDA}}$
        \item $\{a^nb^{n+k}\mid n\geq 0,k\in\{0,1\}\}\in\mathrm{L_{DPDA}}\setminus\mathrm{N_{DPDA}}$
        \item $\{a^nb^{n\cdot 2}\mid n\geq 0\}\in\mathrm{L_{DPDA}}$
        \item $\{a^nb^{n\cdot k}\mid n\geq 0,k\in\{1,2\}\}\in\mathrm{CFL}\setminus\mathrm{L_{DPDA}}$
    \end{enumerate}
\end{problem}
    

\section*{Extra Practice and Thinking}


\begin{problem}[Closure Properties]
    Are (I) $\mathrm{CFL}$, (II) $\mathrm{L_{DPDA}}$, (III) $\mathrm{N_{DPDA}}$ closed under the following operations? Prove or disprove.
    \begin{multicols}{2}
        \begin{enumerate}[(a)]
            \item Union
            \item Concatenation
            \item Iteration
            \item Reversal
            \item Substitution of a regular language
            \item Inverse homomorphism
            \item Right/left derivation
            \item Union with a regular language
        \end{enumerate}    
    \end{multicols}    
\end{problem}


\begin{problem}[Bonus: Context-sensitive Grammar]
    Let $G=(\{S,A,B,C\},\{a,b,c\},\mathcal P,S)$, where:
    \begin{align*}
        \mathcal P=\{&S\rightarrow aSBC\mid aBC, B\rightarrow BBC,  C\rightarrow CC, CB\rightarrow BC,\\ 
        &aB\rightarrow ab, bB\rightarrow bb, bC\rightarrow bc, cC\rightarrow cc\}
    \end{align*}
    What language does it generate? Is the grammar $G$ context-sensitive? If not, find an equivalent context-free grammar.
\end{problem}


\end{document}
