\documentclass[a4paper,12pt]{amsart}

\input{header.tex}

\begin{document}

\thispagestyle{empty}

\section*{NTIN071 A\&G: Tutorial 3 -- Equivalent and minimal representations, testing properties, nondeterminism, subset construction}

\medskip

\subsection*{Teaching goals:} The student is able to

    \begin{itemize}\setlength{\itemsep}{0pt}
        \item define reachable states, state equivalence, reduced automaton, automata homomorphism, NFA, $\epsilon$-NFA
        \item apply the reachable states and state equivalence algorithms to reduce a DFA
        \item design efficient algorithms to test basic properties of automata
        \item apply the subset construction to convert an NFA or $\epsilon$-NFA to a DFA
    \end{itemize}


\section*{In-class problems}


\medskip\begin{problem}[Equivalent and minimal representations]

    For the automata below:

    \begin{enumerate}[(a)]\setlength\itemsep{6pt}
        \item Find and remove all unreachable states.
        \item Determine the state equivalence (indistinguishability) relations. (Moreover, for any distinguishable pair of states find all minimal-length distinguishing words.)
        \item Construct their reducts.        
    \end{enumerate}
    
    \begin{multicols}{2}

        \centering
    
        \begin{tabular}{ r | c c }
            A & a & b \\ \hline
            $\to\ast$ 0 & 1 & 2 \\  
            1 & 3 & 0 \\
            2 & 4 & 5 \\
            3 & 0 & 2 \\
            4 & 2 & 5 \\
            5 & 0 & 3
        \end{tabular}
            
        \begin{tabular}{ r | c c }
            B  & a & b \\ \hline
            $\to\ast$ 0 & 0 & 5 \\  
            1 & 1 & 3 \\
            2 & 2 & 7 \\
            3 & 3 & 2 \\
            $\ast$ 4 & 6 & 1 \\
            5 & 5 & 1 \\
            $\ast$ 6 & 4 & 2 \\
            7 & 7 & 0
        \end{tabular}

    \end{multicols}

\end{problem}


\begin{problem}[Testing properties]

    Consider finite automata $A,B$. Design an algorithm that decides the following properties. (Can you estimate the time complexity?)
    
    \begin{multicols}{4}
    
        \begin{enumerate}[(a)]\setlength\itemsep{6pt}
            \item $L(A)=\emptyset$,
            \item $L(A)=L(B)$,
            \item $L(A)\subseteq L(B)$,
            \item $L(A)$ is finite.
        \end{enumerate}

    \end{multicols}

    Apply the algorithm to the automata $A,B$ from the previous problem.

\end{problem}


\begin{problem}[Subset construction]

    Given a nondeterministic finite automaton with  $\epsilon$-transitions, construct an equivalent reduced DFA.
    
    \begin{center}
        \begin{tabular}{ r | c c c }
            & a & b & $\epsilon$ \\ \hline
            $\to A$ & $\{E\}$ & $\{B\}$ & $\emptyset$ \\
            $B$ & $\emptyset$ & $\{C\}$ & $\{D\}$ \\
            $\to C$ & $\emptyset$ & $\{D\}$ & $\emptyset$ \\
            $\ast D$ & $\emptyset$ & $\emptyset$ & $\emptyset$ \\
            $E$ & $\{F\}$ & $\emptyset$ & $\{B,C\}$\\
            $F$ & $\{D\}$ & $\emptyset$ & $\emptyset$
        \end{tabular}
    \end{center}

\end{problem}



\section*{Extra Practice and Thinking}


\medskip\begin{problem}[Reducing a DFA]    
    
    Reduce the following DFA:
    
    \begin{center}
        \begin{tabular}{ r | c c }
            C & a & b \\ \hline
            $\to$ 1 & 2 & 3 \\
            2 & 2 & 4 \\
            $\ast$ 3 & 3 & 5 \\
            4 & 2 & 7 \\
            $\ast$ 5 & 6 & 3 \\
            $\ast$ 6 & 6 & 6 \\
            7 & 7 & 4 \\
            8 & 2 & 3 \\
            9 & 9 & 4
        \end{tabular}
    \end{center}

\end{problem}


\medskip\begin{problem}[Subset construction]

    Construct an equivalent reduced DFA.    
    
    \begin{center}
        \begin{tabular}{ r | c c c }
            & a & b & $\epsilon$ \\ \hline
            $\ast A$ & $\{A,C\}$ & $\{B\}$ & $\emptyset$ \\
            $B$ & $\{B,D\}$ & $\emptyset$ & $\emptyset$ \\
            $\ast C$ & $\{E\}$ & $\{D\}$ & $\emptyset$ \\
            $D$ & $\{A\}$ & $\{C,D\}$ & $\emptyset$ \\
            $\to\ast E$ & $\emptyset$ & $\emptyset$ & $\{A,C\}$
        \end{tabular}
    \end{center}

\end{problem}


\medskip\begin{problem}[Automata homomorphism] 
    
    Find DFA $A,B$ such that:
    
    \medskip
    
    \begin{enumerate}[(a)]\setlength\itemsep{6pt}
        \item Both are reduced and they are not isomorphic. 
        \item $A$ is homomorphic onto $B$ but they are not isomorphic.
        \item They are equivalent but not isomorphic.
        \item They are both homomorphic onto $C$ but not isomorphic with it. Moreover, $A$ is not homomorphic onto $B$ and $B$ is not homomorphic onto $A$.

        $$
        C=(\{p,q\},\{0,1\},\{((p,0),q),((p,1),p),((q,0),p),((q,1),q)\},p,\{q\})
        $$
    \end{enumerate}

\end{problem}


\medskip\begin{problem}[Regular? Reduce]
    
    Let $L$ be a language over the alphabet $\{a,b\}$ consisting of words that do not contain three consecutive occurrences of the same letter. Decide if $L$ is regular. If it is, find a reduced DFA accepting $L$.

\end{problem}


\medskip\begin{problem}[Subset construction reduced]
    
    Is the result of the subset construction (where we only generate reachable states) always a reduced DFA? Prove or disprove.

\end{problem}


\end{document}
