\documentclass[a4paper,12pt]{amsart}

\input{header.tex}

\begin{document}

\thispagestyle{empty}

\section*{NTIN071 A\&G: Tutorial 9 -- Pushdown automata, conversion between PDA and CFG}

\medskip

\subsection*{Teaching goals:} The student is able to

    \begin{itemize}\setlength{\itemsep}{0pt}
        \item state the formal definition of a PDA, acceptance by empty stack and by final state
        \item construct a pushdown automaton for a given language
        \item convert between acceptance by empty stack and acceptance by final state
        \item convert a context-free grammar into a pushdown automaton
        \item convert a pushdown automaton into a context-free grammar
    \end{itemize}


\section*{In-class problems}


\medskip\begin{problem}[PDA Construction]\label{problem:construct-pda}

    Design a pushdown automaton recognizing the given language. For (a), (b), (c), accept by empty stack; for (d), (e), (f) by final state.

    \medskip

    \begin{enumerate}[(a)]\setlength\itemsep{3pt}
        \item $L=\{ww^R\mid w\in \{0,1\}^*\}$
        \item $L=\{w\in\{(,)\}^*\mid w\text{ is correctly parenthesized}\}$
        \item $L=\{a^ib^jc^k\mid i=j \text{ or } j=k\} $
        \item $L=\{a^{2n}b^{3n}\mid n\geq 0\}$
        \item $L=\{w\in \{0,1\}^*\mid  |w|_0=|w|_1\} $
        \item $L=\{u2v\mid u,v\in \{0,1\}^*\text{ and }|u|\neq |v|\} $
    \end{enumerate}

\end{problem}


\medskip\begin{problem}[Final State vs. Empty Stack]

    Convert selected pushdown automata constructed in the previous problem from final state acceptance to empty stack acceptance, and vice versa. (Try both conversions.)

\end{problem}


\medskip\begin{problem}[CFG to PDA Conversion]

    For a given grammar $G$, construct a PDA $A$ such that $L(G)=N(A)$. Additionally, for a given word $w\in L(G)$, find the leftmost derivation from $G$ and simulate the computation of $A$ (write the accepting sequence of configurations).

    \medskip

    \begin{enumerate}[(a)]\setlength\itemsep{6pt}
        \item $G=(\{S,T,X\},\{a,b\},\mathcal P,S)$ with the following rules, $w=aaaabbb$
            \begin{align*}
        \mathcal P=\{S&\rightarrow aTXb, \\
        T&\rightarrow XTS\mid \epsilon,\\ 
        X&\rightarrow a\mid b\}
        \end{align*}
        \item $G=(\{S,T,X\},\{(,),*,+,1\},P,S)$ with the following rules, $w=1+1*(1+1)$
            \begin{align*}
        P=\{S&\rightarrow S+T\mid T, \\
        T&\rightarrow T*X\mid X,\\ 
        X&\rightarrow 1\mid (S)\}
        \end{align*}
    \end{enumerate}

\end{problem}
    

\medskip\begin{problem}[PDA to CFG Conversion]

    Convert the pushdown automata from Problem 1 (a), (b) into context-free grammars. For a reasonably long word $w$ accepted by the given automaton, find the leftmost derivation of this word in the constructed grammar.

\end{problem}


\section*{Extra Practice and Thinking}


\medskip\begin{problem}[Bonus: Context-Sensitive Grammar]
    
    Consider $G=(\{S,A,B,C\},\{a,b,c\},S,P)$, where:
    \begin{align*}
        P=\{&S\rightarrow aSBC\mid aBC, B\rightarrow BBC,  C\rightarrow CC, CB\rightarrow BC,\\ 
        &aB\rightarrow ab, bB\rightarrow bb, bC\rightarrow bc, cC\rightarrow cc\}
    \end{align*}
    What language does it generate? Is grammar $G$ context-sensitive? If not, find an equivalent context-sensitive grammar.
    
\end{problem}


\medskip\begin{problem}[PDA Construction]

    Design pushdown automata for the following languages. (They may accept by final state or empty stack; for some, construct both; for others, try converting between these two acceptance methods.)

    \medskip

    \begin{enumerate}[(a)]\setlength\itemsep{9pt}
        \item $L=\{w\mid w\in\{0,1\}^*,|w|_1\geq 3\}$        
        \item $L=\{w\in \{0,1\}^*\mid w=w^R\}$        
        \item $L=\{a^ib^jc^k\mid i+j=k\}$        
        \item $L=\{w\in\{(,),[,]\}^*\mid w\text{ is correctly parenthesized}\}$
    \end{enumerate}

\end{problem}


\end{document}
